Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-10-01 paper/current/main.tex
edit first main para

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Construction of affiliation trajectories from publication records}
        \textbf{a.}
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
        A unique identifier is assigned to each each organization and they are assembled into an affiliation trajectory ordered by year of publication.
        \textbf{b.}
        If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

@@

The embedding distance correlates with and can be used to predict real-wold mobility flows better than geographic distance. 
The amount of variance in the expected flux explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using geographic distance and embedding distance as parameters.
The embedding distance produces better predictions of actual flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c-d), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
The embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Our results unequivocally demonstrate that the embedding distance better captures patterns of actual scientific mobility than does geographic distance.

This approach is generalizable to other domains, such as learning a functional distance between airports. 
We learned an embedding of U.S. airports following a similar procedure as scientific mobility, using trajectories formed by the flight itineraries of individual passengers (see Data and Methods).
As with scientific mobility, the embedding approach better predicts real-world flows of airline passengers than does geographic distance. 
Embedding distance explains more of the expected flux between U.S. airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}e) than does geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}f). 
Fitting the data to the gravity model given by Eq.~\ref{eq:gravity_basic} using each of embedding and geographic distance as parameters, and we find that predictions made using embedding distance out-performed those made using geographic distnace (Fig.~\ref{fig:gravity_performance}g-h). 
These results demonstrate that embeddings are applicable to a diverse range of mobility trajectory data.

>>>>

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
    \centering
    \includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
    \caption{
        \textbf{Construction of affiliation trajectories from publication records}
        \textbf{a.}
        An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
        A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication.
        \textbf{b.}
        If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
    }
    \label{fig:methods:mobility_sentence}
\end{figure}

@@

We show that the embedding distance, compared with the geographical distances, predicts actual mobility flows. 
For the case of scientific mobility, the amount of variance in the expected flux explained by the embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
We fit the gravity model given by Eq.~\ref{eq:gravity_basic} (see Data and Methods) and computed the expected flux  $\hat{T}_{ij}$ using the geographic distance and the embedding distance as parameters.
The embedding distance produces better predictions of actual flux than does the geographic distance (Fig.~\ref{fig:gravity_performance}c-d), which also holds for the subsets of only domestic and only international mobility (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
The embedding distance defined from the cosine distance between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion-based'' network distance measures such as the personalized-page rank scores (see \SI) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}). These results are consistent in the U.S. flight itinerary data.
As with scientific mobility, the embedding approach better predicts actual flows of airline passengers than the geographic distance. 
The embedding distance explains more of the expected flux between U.S. airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}f). 
By fitting the gravity model given by Eq.~\ref{eq:gravity_basic} using each of the embedding distance and the geographic distance as parameters, and we find that predictions with embedding distance out-performed those with the geographic distance (Fig.~\ref{fig:gravity_performance}g-h). 
Our results demonstrate that the embedding distance better captures patterns of actual mobility than does the geographic distance across datasets.
